# osmdatar

R package for downloading OSM data. Current status (on test data of highways
only):

method    |   Computation time (s)
------  | ------    
osmplotr    |   1.86
hrbrmstr    |   1.47    
Rcpp    |   0.25


------

## Install

```{r, message=FALSE}
Sys.setenv ('PKG_CXXFLAGS'='-std=c++11')
setwd ("..")
#devtools::document ("osmdatar")
devtools::load_all ("osmdatar")
setwd ("./osmdatar")
Sys.unsetenv ('PKG_CXXFLAGS')
```

-----

## Speed comparisons

The `osmplotr` package uses `XML` to process the API query, and `osmar` to
convert the result to `sp` structures. The
[overpass](https://github.com/hrbrmstr/overpass/) repo of
[hrbrmstr](https://github.com/hrbrmstr) uses `xml2` and thus all cribbed
functions are called `_xml2_`. All of my new `Rcpp` functions described below
are then appended with `_3`.  In all cases, the `process_xml_...` functions
convert an xml document to the same SpatialLinesDataFrame of highways.

```{r, message=FALSE}
library (microbenchmark)
```

## 1. Osmar / osmplotr

```{r, eval=FALSE}
bbox <- matrix (c (-0.12, 51.51, -0.11, 51.52), nrow=2, ncol=2) 
doc <- get_xml_doc (bbox=bbox)
mb <- microbenchmark ( obj <- process_xml_doc (doc), times=100L)
tt <- formatC (mean (mb$time) / 1e9, format="f", digits=2)
```
```{r, echo=FALSE}
tt <- 1.86
```
```{r}
cat ("Mean time to convert with osmar =", tt, "s\n")
```

## 2. hrbrmstr

```{r, eval=FALSE}
doc2 <- get_xml2_doc (bbox=bbox)
mb2 <- microbenchmark ( obj2 <- process_xml2_doc (doc2), times=100L )
tt2 <- formatC (mean (mb2$time) / 1e9, format="f", digits=2)
```
```{r, echo=FALSE}
tt2 <- 1.47
```
```{r}
cat ("Mean time to convert with hrbrmstr code =", tt2, "\n")
```

The code of **hrbrmstr** using `dplyr` is (only) around 20% faster than using
`osmar`. And then for the C++ version ...

## 3. Rcpp

The function calls *should* compile properly within the `load_all` call, but in
case they don't they can be loaded manually here:
```{r, eval=FALSE}
Sys.setenv ('PKG_CXXFLAGS'='-std=c++11')
Rcpp::sourceCpp('src/get_highways.cpp')
Sys.unsetenv ('PKG_CXXFLAGS')
```
(One reason this might be necesssary is because `devtools::document` fails to
insert the necessary line `useDynLib(osmdatar)` in `NAMESPACE`. Re-inserting
this manually should fix the problem.)

And then the actual tests, performed here with 4 versions, the first of which
uses the `Rcpp` function `get_highways`, which returns a list of matrices of 2
columns.  The subsequent 3 use `get_highways_with_id`, which returns a list of
matrices of 3 columns, the first of which contains the OSM way ID.

The differences between the corresponding `R` functions are:

1. `process_xml_doc3a` uses a simple loop.
2. `process_xml_doc3b` uses a simple loop (and so time differences arises only
   through the third ID column from Rcpp)
3. `process_xml_doc3c` is same as 2/b, but uses `lapply`
4. `process_xml_doc3d` uses the `dplyr::groupby` approach of **hrbrmstr**

```{r, eval=FALSE}
txt <- get_xml_doc3 (bbox=bbox)
mb3a <- microbenchmark ( obj3 <- process_xml_doc3a (txt), times=100L )
mb3b <- microbenchmark ( obj3 <- process_xml_doc3b (txt), times=100L )
mb3c <- microbenchmark ( obj3 <- process_xml_doc3c (txt), times=100L )
mb3d <- microbenchmark ( obj3 <- process_xml_doc3d (txt), times=100L )
tt3a <- formatC (mean (mb3a$time) / 1e9, format="f", digits=2)
tt3b <- formatC (mean (mb3b$time) / 1e9, format="f", digits=2)
tt3c <- formatC (mean (mb3c$time) / 1e9, format="f", digits=2)
tt3d <- formatC (mean (mb3d$time) / 1e9, format="f", digits=2)
```
```{r, echo=FALSE}
tt3a <- 0.25
tt3b <- 0.25
tt3c <- 0.25
tt3d <- 0.35
```
```{r}
cat ("Mean times to convert with Rcpp code = (", 
     tt3a, ", ", tt3b, ", ", tt3c, ", ", tt3d, ")\n", sep="")
```

The `dplyr` approach of **hrbrmstr** is thus considerably slower than simply
using direct loops, while the addition of an extra ID column to the `Rcpp`
output does not slow things down at all. It is possible that the `Rcpp` routine
could be sped up a little, but this is unlikely to make any difference compared
to the necessity of looping / `dplyr`-ing within `R`. The slowness simply comes
from the fact that so-called `Spatial___DataFrame` structures are actually
**lists**, yet do not redefine any assignment operators for list elements, and
so inevitably require loops.
