# osmdatar
R package for downloading OSM data

------

```{r, echo=FALSE, message=FALSE}
setwd ("..")
devtools::load_all ("osmdatar")
setwd ("./osmdatar")
```

## Speed comparisons

The `osmplotr` package uses `XML` to process the API query, and `osmar` to
convert the result to `sp` structures. The
[overpass](https://github.com/hrbrmstr/overpass/) repo of 
[hrbrmstr](https://github.com/hrbrmstr) uses `xml2` and thus all cribbed
functions are called `_xml2_`.

```{r, message=FALSE}
library (microbenchmark)
```

The **examples** code is repeated here with `microbenchmark` comparisons

```{r, eval=FALSE}
bbox <- matrix (c (-0.12, 51.51, -0.11, 51.52), nrow=2, ncol=2)
doc <- get_xml_doc (bbox=bbox)
mb <- microbenchmark ( obj <- process_xml_doc (doc), times=100L)
tt <- formatC (mean (mb$time) / 1e9, format="f", digits=2)
```
```{r, echo=FALSE}
tt <- 2.05
```
```{r}
cat ("Mean time to convert with osmar =", tt, "s\n")
```

```{r, eval=FALSE}
doc2 <- get_xml2_doc (bbox=bbox)
mb2 <- microbenchmark ( obj2 <- process_xml2_doc (doc2), times=100L )
tt2 <- formatC (mean (mb2$time) / 1e9, format="f", digits=2)
```
```{r, echo=FALSE}
tt2 <- 1.48
```
```{r}
cat ("Mean time to convert with hrbrmstr code =", tt2, "\n")
```

The code of **hrbrmstr** using `dplyr` is (only) around 30% faster than using
`osmar`. Definitely time to try a C++ version ...
