---
title: "3. Translation of OSM to Simple Features"
author: "Mark Padgham"
date: "`r Sys.Date()`"
output:
    html_document:
        toc: true
        toc_float: true
        theme: flatly
vignette: >
  %\VignetteIndexEntry{3. OSM to Simple Features}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## 1. OpenStreetMap Data Structure

[OpenStreetMap (OSM)](https://www.openstreetmap.org/) data has a unique structure that
is not directly reconcilable with other modes of representing spatial data,
notably including the widely-adopted 
[Simple Features (SF)](https://www.ogc.org/standards/sfa) 
scheme of the 
[Open Geospatial Consortium (OGC)](https://www.ogc.org). 
The three primary spatial objects of OSM are:

1. `nodes`, which are directly translatable to spatial points

2. `ways`, which may be closed, in which case they form polygons, or unclosed,
   in which case they are (non-polygonal) lines.

3. `relations` which are higher-level objects used to specify relationships
   between collections of ways and nodes. While there are several recognised
   categories of `relations`, in spatial terms these may be reduced to a binary
   distinction between:

   - `multipolygon` relations, which specify relationships between an
   exterior polygon (through designating `role='outer'`) and possible inner
   polygons (`role='inner'`). These may or may not be designated with
   `type=multipolygon`. Political boundaries, for example, often have
   `type=boundary` rather than explicit `type-multipolygon`. `osmdata`
   identifies multipolygons as those `relation` objects having at least one
   member with `role=outer` or `role=inner`.

   - In the absence of `inner` and `outer` roles, an OSM relation is assumed to
   be non-polygonal, and to instead form a collection of non-enclosing lines.

--------

## 2. Simple Features Data Structure

The representation of spatial objects as Simple Features is 
[described at length by the OGC](https://www.ogc.org/standards/sfo),
with this document merely reviewing relevant aspects. The SF system assumes that
spatial features can be represented in one of seven distinct primary classes,
which by convention are referred to in all capital letters.  Relevant classes
for OSM data are:

1. POINT
2. MULTIPOINT
3. LINESTRING
4. MULTILINESTRING
5. POLYGON
6. MULTIPOLYGON

(The seventh primary class is `GEOMETRYCOLLECTION`, which contains several
objects with different geometries.) An SF (where that acronym may connote both
singular and plural) consists of a sequence of spatial coordinates, which for
OSM data are only ever `XY` coordinates represented as strings enclosed within
brackets.  In addition to coordinate data and associated coordinate reference
systems, an SF may include any number of additional data which quantify or
qualify the feature of interest.  In the 
[`sf` extension to `R`](https://cran.r-project.org/package=sf), for example, a
single SF is represented by one row of a `data.frame`, with the geometry stored
in a single column, and any number of other columns containing these additional
data.

Simple Feature geometries are referred to in this vignette using all capital
letters (such as `POLYGON`), while OSM geometries use lower case (such as
`polygon`). Similarly, the Simple Features standard of the OGC is referred to as
`SF`, while the `R` package of the same name is referred to as `R::sf`--upper
case `R` followed by lower case `sf`. Much functionality of `R::sf` is
determined by the underlying 
[Geospatial Data Abstraction Library](https://gdal.org) (`GDAL`; described
below). Representations of data are often discussed here with reference to
`GDAL/sf`, in which case it may always be assumed that the translation and
representation of data are determined by `GDAL` and not directly by the creators
of `R::sf`.


--------

## 3. How `osmdata` translates OSM into Simple Features

### 3.1. OSM Nodes

OSM nodes translate directly into `SF::POINT` objects, with all OSM `key-value`
pairs stored in additional `data.frame` columns.

### 3.2. OSM Ways

OSM ways may be either polygons or (non-polygonal) lines. `osmdata` translates
these into `SF::LINESTRING` and `SF::POLYGON` objects, respectively. Although
polygonal and non-polygonal ways may have systematically different `key` fields,
they are conflated here to the single set of `key` values common to all `way`
objects regardless of shape. This enables direct comparison and uniform
operation on both `SF::LINESTRING` and `SF::POLYGON` objects.

### 3.3 OSM Relations

OSM relations comprising members with `role=outer` or `role=inner` are
translated into `SF::MULTIPOLYGON` objects; otherwise they form
`SF::MULTILINESTRING` objects. As in the preceding case of OSM ways, potentially
systematic differences between OSM `key` fields for `multipolygon` and other
`relation` objects are ignored in favour of returning identical `key` fields in
both cases, whether or not `value` fields for those `key`s exist.

#### 3.3(a) Multipolygon Relations

An OSM multipolygon is translated by `osmdata` into a single `SF::MULTIPOLYGON`
object which has an additional column specifying `num_members`. The `SF`
geometry thus consists of a list (an `R::List` object) of this number of
polygons, the first of which is the `outer` polygon, with all subsequent members
forming closed inner rings (either individually or in combination).

Each of these inner polygons are also represented as one or more OSM objects, which
will generally include detailed data on the individual components **not** able
to be represented in the single multipolygon representation.  Each inner polygon
is therefore additionally stored in the `sf::MULTIPOLYGON` `data.frame` along
with all associated data.  Thus the row containing a multipolygon of
`num_polygon` polygons is followed by `num_polygon - 1` rows containing the data
for each `inner` polygon.

Note that OSM `relation` objects generally have fewer (or different) `key-value`
pairs than do OSM `way` objects. In the OSM system, data describing the detailed
properties of the constituent ways of a given OSM relation are stored with those
`ways` rather than with the `relation`. `osmdata` follows this general
principle, and stored the geometry of all `ways` of a `relation` with the
relation itself (that is, as part of the `MULTIPOLYGON` or `MULTILINESTRING`
object), while those ways are also stored themselves as `LINESTRING` (or
potentially `POLYGON`) objects, from where their additional `key-value` data may
be accessed.


#### 3.3(b) Multilinestring Relations

OSM `relations` that are not `multipolygons` are translated into
`SF::MULTILINESTRING` objects. Each member of any OSM relation is attributed a
`role`, which may be empty. `osmdata` collates all ways within a relation
according to their `role` attributes. Thus, unlike multipolygon relations which
are always translated into a single `sf::MULTIPOLYGON` object, multilinestring
relations are translated by `omsdata` into potentially several
`sf::MULTILINESTRING` objects, one for each unique role.

This is particularly useful because `relations` are often used to designated
extended `highways` (for example, designated bicycle routes or motorways), yet
these often exist in `primary` and `alternative` forms, with these categories
specified in roles. Separating these roles enables ready access to any desired
role.

These multilinestring objects also have a column specifying `num_members`, as
for multipolygons, with the primary member followed by `num_members` rows, one
for each member of the multilinestring.

---------

## 4. `GDAL` Translation of OSM into Simple Features

The `R` package [`sf`](https://cran.r-project.org/package=sf) provide an `R`
implementation of Spatial Features, and provides a wrapper around GDAL for
reading geospatial data. `GDAL` provides a ['driver' to read OSM
data](https://gdal.org/drv_osm.html), and thus `sf` can also be used to read
`OSM` data in `R`, 
[as detailed in the main `osmdata` vignette](https://docs.ropensci.org/osmdata/articles/osmdata.html).
However, the `GDAL` translation of OSM data differs in several important ways
from the `osmdata` translation.

The primary difference is that GDAL only returns *unique* objects of each
spatial (SF) type. Thus `sf::POINT` objects consist of only those points that
are not otherwise members of some 'higher' object (`line`, `polygon`, or
`relation` objects).  Although a given set of OSM data may actually contain a
great many points, attempting to load these with
```{r, eval=FALSE}
sf::st_read (file, layer = 'points')
```
will generally return surprisingly few points.

### 4.1. OSM Nodes

Apart from the numerical difference arising through `osmdata` returning an
`sf::POINTS` structure containing **all** nodes within a given set of OSM data,
while `sf::st_read (file, layer='points')` returns only those points not
represented in other structure, the representation of points remains otherwise
broadly similar. The only other major difference is that `osmdata` retains all
`key-value` pairs present in a given set of OSM data, whereas `GDAL/sf` only
retains a select few of these. Moreover, the `keys` returned by `GDAL/sf` are
pre-defined and invariant, meaning that data returned from `sf::st_read (...)`
may often contain `key` columns in the resultant `data.frame` which contain no
(non-`NA`) data. This difference is illustrated in an example repeated here from
the  
[main `osmdata` vignette](https://docs.ropensci.org/osmdata/articles/osmdata.html),
with the same principles applying to all of the following classes of OSM data.

The following three lines define a query and download the resultant data to an
`XML` file.
```{r trentham, eval=FALSE}
q <- opq (bbox = 'Trentham, Australia')
q <- add_osm_feature (q, key = 'name') # any named objects
osmdata_xml (q, 'trentham.osm')
```
These data may then be converted into SF representations using either `R::sf` or
`osmdata`, with OSM `keys` being the column names of the resultant `data.frame`
objects.
```{r, eval=FALSE}
names (sf::st_read ('trentham.osm', layer = 'points', quiet = TRUE))
```
```{r, echo=FALSE}
c ("osm_id",     "name",       "barrier",    "highway",    "ref",
"address",    "is_in",      "place",      "man_made",   "other_tags",
"geometry")
```
```{r, eval=FALSE}
names (osmdata_sf (q, 'trentham.osm')$osm_points)
```
```{r, echo=FALSE}
c ("osm_id",           "name",             "X_description_",   "X_waypoint_",
"addr.city",        "addr.housenumber", "addr.postcode",    "addr.street",
"amenity",          "barrier",          "denomination",     "foot",
"ford",             "highway",          "leisure",          "note_1",
"phone",            "place",            "railway",  "railway.historic",
"ref",              "religion",         "shop",             "source",
"tourism",          "waterway",         "geometry")
```
`osmdata` returns far more `key` fields than does `GDAL/sf`. More importantly,
however, `GDAL/sf` returns pre-defined `key` fields regardless of whether they
contain any data:
```{r, eval=FALSE}
addr <- sf::st_read ('trentham.osm', layer = 'points', quiet = TRUE)$address
all (is.na (addr))
```
```{r, echo=FALSE}
TRUE
```
In contrast, `osmdata` returns only those `key` fields which contain data (and
so excludes `address` in the above example).


### 4.2. OSM Ways

As for points, `GDAL/sf` only returns those ways that are not represented or
contained in 'higher' objects (OSM relations interpreted as `SF::MULTIPOLYGON`
or `SF::MULTILINESTRING` objects). `osmdata` returns all ways, and thus enables,
for example, examination of the full attributes of any member of a multigeometry
object. This is not possible with the `GDAL/sf` translation.  As for points, the
only additional difference between `osmdata` and `GDAL/sf` is that `osmdata`
retains all `key-value` pairs, whereas `GDAL` retains only a select few.

### 4.3 OSM Relations

Translation of OSM relations into Simple Features differs more significantly
between `osmdata` and `GDAL/sf`.

#### 4.3(a) Multipolygon Relations

As indicated above, multipolygon relations are translated in broadly comparable
ways by both `osmdata` and `sf/GDAL`. Note, however, the `way` members of an OSM
relation may be specified in arbitrary order, and the multipolygonal way may not
necessarily be traced through simply following the segments in the order
returned by `sf/GDAL`.

#### 4.3(b) Multilinestring Relations

Linestring relations are simply read by GDAL directly in terms of the their
constituent ways, resulting in a single `SF::MULTILINESTRING` object that
contains exactly the same number of lines as the ways in the OSM relation,
regardless of their `role` attributes. Note that `roles` are frequently used to
specify `alternative` multi-way routes through a single OSM relation. Such
distinctions between primary and alternative are erased with `GDAL/sf` reading.

## 5 Examples

### 5.1 Routing

Navigable paths, routes, and ways are all tagged within OSM as `highway`,
readily enabling an `overpass` query to return only `ways` that can be used for
routing purposes. Routes are nevertheless commonly assembled within OSM
relations, particularly where they form major, designated transport ways such as
long-distance foot or bicycle paths or major motorways.

#### 5.1(a) Routing with `sf/GDAL`

A query for `key=highway` translated through `GDAL/sf` will return those ways
not part of any 'higher' structure as `SF::LINESTRING` objects, but components
of an entire transport network might also be returned as:

1. `SF::MULTIPOLYGON` objects, holding all single ways which form simple
   polygons (that is, in which start and end points are the same); 
2. `SF::MULTIPOLYGON` objects holding all single (non-polygonal) ways which
   combine to form an `OSM multipolygon` relation (that is, in which the
   collection of ways ultimately forms a closed `role=outer` polygon).
3. `SF::MULTILINESTRING` objects holding all single (non-polygonal) ways which
   combine to form an OSM relation that is not a multipolygon.

Translating these data into a single form usable for routing purposes is not
simple. A particular problem that is extremely difficult to resolve is
reconciling the `SF::MULTIPOLYGON` objects with the geometry of the
`SF::LINESTRING` objects. Highway components contained in `SF::MULTIPOLYGON`
objects need to be re-connected with the network represented by the
`SF::LINESTRING` objects, yet the OSM identifiers of the `MULTIPOLYGON`
components are removed by `sf/GDAL`, preventing these components from being
directly re-connected. The only way to ensure connection would be to re-connect
those geographic points sharing identical coordinates. This would require code
too long and complicated to be worthwhile demonstrating here.

#### 5.1(b) Routing with `osmdata`

`osmdata` retains all of the underlying ways of 'higher' structures
(`SF::MULTIPOLYGON` or `SF::MULTILINESTRING` objects) as 
`SF::LINESTRING` or `SF::POLYGON` objects. The geometries of the latter objects
duplicate those of the 'higher' relations, yet contain additional `key-value`
pairs corresponding to each way. Most importantly, the OSM ID values for all
members of a `relation` are stored within that relation, readily enabling the
individual ways (`LINESTRING` or `POLYGON` objects) to be identified from the
`relation` (`MULTIPOLYGON` or `MULTILINESTRING` object).

The `osmdata` translation thus readily enables a singularly complete network to
be reconstructed by simply combining the `SF::LINESTRING` layer with the
`SF::POLYGON` layer. These layers will always contain entirely independent
members, and so will always be able to be directly combined without duplicating
any objects.
