---
title: "1. osmdata"
author: "Mark Padgham"
date: "`r Sys.Date()`"
output:
    html_document:
        toc: true
        theme: flatly
vignette: >
  %\VignetteIndexEntry{1. osmdata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## 1. A quick introduction

The development version of the `osmdata` R package can be installed with the `devtools` (which can be installed with `installed.packages("devtools")`) as follows:

```{r install, eval=FALSE}
devtools::install_github("osmdatar/osmdata")
```

Once installed, it can be loaded in the usual way:

```{r}
library(osmdata)
```

`osmdata` uses the [`overpass` API](http://overpass-api.de) to download
OpenStreetMap (OSM) data and convert the results to either [Simple Features
(SF)](https://cran.r-project.org/package=sf) or
[`Spatial`](http://cran.r-project.org/package=sp) objects, defined by the `sf` and `sp` packages respectively. `overpass` is a C++ library that serves OSM data over the web (see [github.com/drolbr/Overpass-API](https://github.com/drolbr/Overpass-API) for its source code.
It has its
own query language, with all queries extracting data within a defined bounding
box. A query may be initiated using the `osmdata` function `opq()`, which
accepts the single argument of a bounding box.

```{r opq1}
q <- opq (bbox=c(51.1,0.1,51.2,0.2))
```
Bounding boxes may also be defined by simply passing the name of a desired area,
```{r opq2}
q <- opq (bbox="Greater London, U.K.")
identical (opq (bbox="Greater London, U.K."), opq (bbox="greater london uk"))
```
An `osmdata` query consists of an initial `opq()` query and one or more features:
```{r opq3}
q <- opq (bbox="greater london uk")
q <- add_feature (q, key="highway", value="motorway")
```
The corresponding data may then be extracted as `R Spatial` objects with,
```{r osmdata_sp}
osmdata_sp (q)
```
or an `R Sinple Features` objects with
```{r osmdata_sf}
osmdata_sf (q)
```

`overpass` data are returned in OSM XML format, and may be directly returned with
```{r osmdata_xml}
doc <- osmdata_xml (q)
```
The function `osmdata_xml()` also allows data to be saved and used for
subsequent queries through specifying a filename.
```{r osmdata_xml2, eval=TRUE}
osmdata_xml (q, "motorways.osm")
```
These downloaded data can then be directly processed by simply passing the same
filename to `osmdata_sf/sp`
```{r osmdata_spsf, eval=FALSE}
dat <- osmdata_sp (q, "motorways.osm")
dat <- osmdata_sf (q, "motorways.osm")
```
Note that the query, `q`, does not have to be submitted, but
```{r osmdata_spsf2, eval=TRUE}
dat <- osmdata_sf (doc="motorways.osm")
```




## 2. The overpass API

The `overpass` API is ...

### 2.1. overpass queries

`overpass` has its own query language which largely works by extracting data
based on specific combinations of `key-value` pairs, with `osmdata`
implementing a simplified version. Full details of the `overpass` query
language are given in the 
[Query Language Guide](http://wiki.openstreetmap.org/wiki/Overpass_API/Overpass_QL) 
(and see also the 
[overpass API Language Guide](http://wiki.openstreetmap.org/wiki/Overpass_API/Language_Guide)
).

As demonstrated above, an `osmdata` query begins by specifying a bounding box
with the function `opq()`, followed by specifying desired OSM features with
`add_feature()`.  
```{r kunming}
q <- opq (bbox="Kunming, China")
q <- add_feature (q, key="natural", value="water")
q <- add_feature (q, key="name:en", value="Dian", exact=FALSE)
```
The parameter `exact=FALSE` controls the two ways of matching `key-value` pairs
in `overpass`: Either exactly and entirely, or through a partial string match.
With the default `exact=TRUE`, this query would only return objects named
"Dian", and not, for example, any named "Lake Dian". Using `exact=FALSE` will
return "Lake Dian" through just specifying `value="Dian"`.

Each successive feature added with `add_feature()` is added to previous
features, so, for example, extending the previous query by an additional
`key-value` pair to 
```{r kunming2}
q <- add_feature (q, key="highway")
```
will request all objects that are both water bodies with names matching "Dian"
**AND** that are also highways. Unsurprisingly, this query returns no data
```{r kunming-sf}
nrow (osmdata_sf (q)$osm_points)
```

Distinct and different features must be requested with distinct and different
queries:
```{r kunming-all, eval=FALSE}
q0 <- opq (bbox="Kunming, China")
q1 <- add_feature (q0, key="natural", value="water")
q1 <- add_feature (q1, key="name:en", value="Dian", exact=FALSE)
dat_water <- osmdata_sf (q1)
q2 <- add_feature (q0, key="highway")
dat_highway <- osmdata_sf (q2)
```

The function `osmdata_sf` returns an object of class `osmdata`, as described in
detail below.

The detailed structure of `overpass` queries can best be appreciated by using
the [interactive online query builder](http://overpass-turbo.eu), which includes a
helpful corrector function for incorrectly formatted queries. An `osmdata` query
includes the two primary fields of `bbox` and `qry`, with the latter passed
directly to the `overpass` API. Any queries constructed with the 
[interactive online query builder](http://overpass-turbo.eu) can also be pasted
directly into the `qry` component of an `osmdata` query.


## 3. `osmdata` queries

As indicated above, `osmdata` queries are generally simplified versions of
potentially more complex `overpass` queries, although arbitrarily complex
`overpass` queries may be pasted directly into the `$qry` field of an `osmdata`
query. Also as illustrated above, `osmdata` queries are generally constructed
through initiating a query with `opq()`, and then specifying OSM features in
terms of `key-value` pairs with `add_feeature()`. 

At its simplest, `add_feature()` requires a single `key`, which by default will
return all objects specifying any value for that `key`: 
```{r opq-london}
q0 <- opq(bbox="London City, U.K.")
q1 <- add_feature (q0, key="highway")
q1 <- add_feature (q1, key="name")
```
Passing that query to `osmdata_xml/sf/sp()` will return all named highways
within the bounding box.

### 3.1. Building a query

### 3.2. Submitting a query

As demonstrated above, `osmdata` queries may be submitted to `overpass` using
one of three functions specified by the kind of data they return.
`osmdata_xml()` returns raw `XML` data, and enables these to be saved directly
to disk for processing with other software.

Start with a new query
```{r london-buildings}
q <- opq (bbox=c(-0.12,51.51,-0.11,51.52)) # Central London, U.K.
q <- add_feature (q, key='building')
```
The OSM data can then be saved as an XML document with:
```{r osmdata_xml3}
osmdata_xml (q, "buildings.osm")
```
The other two functions return data in the two primary `R` formats for storing
and processing spatial data: `R Spatial (sp)` and `R Simple Features (sf)`.
Both of these functions accept optional arguments specifying previously
downloaded files (using `osmdata_xml()`) or an `XML` document.  
```{r osmdata_with_files}
dat <- osmdata_sf (q, "buildings.osm")
doc <- osmdata_xml (q, file="bldgs2.osm")
identical (osmdata_sf (q, doc), osmdata_sf (q, "bldgs2.osm"))
```

## 4. The `osmdata` object

The `osmdata` extraction functions, `osmdata_sf()`, and `osmdata_sp()`, both
return objects of class `osmdata`, the structure of which is shown with the
generic print method as demonstrated above. 
```{r osmdata_with_files2}
class (osmdata_sf (q, "buildings.osm"))
class (osmdata_sp (q, "buildings.osm"))
```
The actual spatial data contained in an `osmdata` object are of either `sp`
format when extracted with `osmdata_sp()` or `sf` format when extracted with
`osmdata_sf()`. 
```{r osmdata_with_files3}
class (osmdata_sf (q, "buildings.osm")$osm_lines)
class (osmdata_sp (q, "buildings.osm")$osm_lines)
```

### 4.1. The `osmdata_xml()` function

`osmdata_xml()` returns OSM data in native XML format, which are stored by
convention with the file suffix `.osm`. This function also allows raw data to
be saved to disk, as demonstrated above. The `XML` data are formatting using
the `R` package `xml2`, and may be processed within `R` using any methods
compatible with such data, or may be processed by any other software able to
load the `XML` data directly from disk.
```{r}
class (osmdata_xml (q))
```

### 4.2. The `osmdata_sf()` function

`osmdata_sf()` returns OSM data in 
[Simple Features (SF)](http://www.opengeospatial.org/standards/sfo) 
format, defined by the
[Open Geospatial Consortium](http://www.opengeospatial.org), and implemented in
the `R` package [`sf`](https://cran.r-proejct.org/package=sf). This package
provides a direct interface to the `C++` 
[Graphical Data Abstraction Library (GDAL)](http://gdal.org) which also includes
a so-called [`driver` for OSM data](http://www.gdal.org/drv_osm.html). This
means that OSM data may also be read directly with `sf`, rather than using
`osmdata`. Data must first be saved to disk, which can for example be readily
achieved through downloading directly from the 
[overpass interactive query builder](http://overpass-turbo.eu). 

Queries constructed in `osmdata` can also be pasted directly into `overpass
turbo`. The following example is based on this query: 
```{r trentham}
q <- opq (bbox="Trentham, Australia")
q <- add_feature (q, key="name") # any named objects
osmdata_xml (q, "trentham.osm")
```


`sf` can then read such data independent of `osmdata` though
```{r sf1}
sf::st_read ("trentham.osm", layer="lines")
```
The `GDAL` drivers used by `sf` can only load single 'layers' of features, for
example, `points`, `lines`, or `polygons`. In contrast, `osmdata` loads all
features simultaneously:
```{r osmdata_sf2}
osmdata_sf (q, "trentham.osm")
```
Even for spatial objects of the same type (the same 'layers' in `sf`
terminology), `osmdata` returns considerably more objects. The raw sizes of data
returned can be compared with

```{r object-sizes}
as.numeric (object.size (osmdata_sf (q, "trentham.osm")$osm_lines) /
    object.size (sf::st_read ("trentham.osm", layer="lines", quiet=TRUE)))
```
And the `osmdata lines` contain over five times as much data.  The primary
difference between `sf/GDAL` and `osmdata` is that the former returns only those
objects unique to each category of spatial object.  Thus OSM nodes (`points` in
`sf/osmdata` representations) include, in `sf/GDAL` representation, only those
points which are not part of any other objects (such as lines or polygons). In
contrast, `osmdata point` objects include all points regardless of whether or
not these are represented in other spatial objects.  Similarly, 'line' objects
in 'sf/GDAL' exclude any lines that are part of other objects such as
'multipolygon' or 'multiline' objects.

This processing of data by `sf/GDAL` has two important implications:

1. An implicit hierarchy of spatial objects is enforced through including
elements of objects only at their 'highest' level of representation, where
`multipolygon` and `multiline` objects are assumed to be at 'higher' levels
than `polyon` or `line` objects, and these in turn are at 'higher' levels than
`point` objects. `osmdata` makes no such hierarchical assumptions.

2. All OSM are structured by giving each object a unique identifier so that the
components of any given object (the nodes of a line, for example, or the lines
of a multipolygon) can be referenced in terms of these identifiers alone. This
enables the full details for any component of a given object to be extracted by
examining that component. The `sf/GDAL` representation obviates this ability
through removing these IDs and reducing everything to geometries alone. This
means, for example, that the `key-value` pairs of the `line` or `polygon`
components of `multipolygon` can never be extracted from an `sf/GDAL`
representation. In contrast, `osmdata` retains all unique identifiers for all
OSM objects, and so readily enables, for example, the properties of all `point`
objects of a `line` to be extracted.

Finally, note that `osmdata` will generally extract OSM data considerably
faster than equivalent `sf/GDAL` routines (as detailed [here](git-wiki)).


### 4.3. The `osmdata_sp()` function

As with `osmdata_sf()` described above, OSM data may be converted to `sp`
format without using `osmdata` via the `sf` functions 
```{r sf_sp, eval=TRUE}
dat <- sf::st_read ("buildings.osm", layer="multipolygons")
dat_sp <- as (dat, "Spatial")
class (dat_sp)
```
These data are extracted using the GDAL, and so suffer all of the same
shortcomings mentioned above. Note differences in the amount of data returned:
```{r}
dim (dat_sp)
dim (osmdata_sf (q, doc='buildings.osm')$osm_polygons); dim (osmdata_sf (q, doc='buildings.osm')$osm_multipolygons)
```

```{r, echo=FALSE}
for (f in list.files (pattern="\\.osm"))
    if (file.exists (f)) file.remove (f)
```

## 5. Recursive searching

OSM data are structured in three hierarchical levels:

1. `nodes` representing spatial points 

2. `ways` representing lines, both as `polygons` (with connected ends) and
   non-polygonal `lines`

3. `relations` representing more complex objects generally comprising
   collections of `ways` and/or `nodes. Examples include `multipolygon
   relations` comprising an outer polygon (which may itself be made of several
   distinct `ways` which ultimately connect to form a single circle), and
   several inner polygons.

`osmdata` contains several functions which can be used to extract the components
of individual objects.

1. `osm_points` extracts all `point` or `node` objects 

2. `osm_lines` extracts all `way` objects that are `lines` (that are, that are
   not `polygons`) 

3. `osm_polygons` extracts all `polygon` objects

4. `osm_multilines` extracts all `multiline` objects; and

5. `osm_multipolygons` extracts all `multipolygon` objects.
